---
title: "Plot phylogenetic trees for sequence capture"
output: html_document
---

This document describes the process of plotting phylogenetic trees for sequence capture output  
The trees need to be in Newick format and should be in the current directory  

Results from IQ-TREE and ASTRAL will have multiple node labels (concordance factors) 
To get the concordance trees into Newick properly, use `concord_to_newick.py`   
Similarly, use `astral_parse.py` to separate node labels for ASTRAL (posterior probabilities and quartet support)  

The outgroup file should have a single tip label per line  
The correspondence between sample IDs and sample labels should be in a tab-separated text file, one line per sample  
```{r}
suppressMessages(library(ape))
suppressMessages(library(phytools))
suppressMessages(library(svglite))
concat_tree <- "concat.treefile"
concat_scf <- "concord_newick_scf.tre"
concat_gcf <- "concord_newick_gcf.tre"
astral_p <- "astral_p.tre"
astral_q <- "astral_q.tre"
outgroup_file <- "outgroup.txt"
samples_file <- "samples.tab"
run_poly <- FALSE
```

If running a polytomy test, add that output
```{r}
run_poly <- TRUE
astral_t <- "astral_poly.tre"
astral_tc <- "astral_poly_collapsed.tre"
```

Read in the samples file
```{r}
if (file.exists(samples_file)) {
	samples_present <- TRUE
	sample_table <- read.table(samples_file, sep = "\t")
	colnames(sample_table) <- c("ID", "label")
} else {
	samples_present <- FALSE
}
```

Read in the outgroups file
```{r}
if (file.exists(outgroup_file)) {
	outgroup_present <- TRUE
	outgroups <- read.table(outgroup_file)[, 1]
} else {
	outgroup_present <- FALSE
}
```

Read in the tree files
```{r}
c_tree <- read.tree(concat_tree)
cs_tree <- read.tree(concat_scf)
cg_tree <- read.tree(concat_gcf)
ap_tree <- read.tree(astral_p)
aq_tree <- read.tree(astral_q)
if (run_poly) {
	at_tree <- read.tree(astral_t)
	atc_tree <- read.tree(astral_tc)
	tree_list <- list(c_tree, cs_tree, cg_tree, ap_tree, aq_tree, at_tree, atc_tree)
} else {
	tree_list <- list(c_tree, cs_tree, cg_tree, ap_tree, aq_tree)
}
```

If there is an outgroup, root the trees
```{r}
if (outgroup_present) {
	for (index in seq_len(length(tree_list))) {
		tree <- tree_list[[index]]
		these_outgroups <- outgroups[outgroups %in% tree$tip.label]
		rootnode <- getMRCA(tree, as.character(these_outgroups))
		position <- 0.5 * tree$edge.length[which(tree$edge[, 2] == rootnode)]
		rooted_tree <- reroot(tree, rootnode, position, edgelabel = TRUE)
		tree_list[[index]] <- rooted_tree
	}
}
```

If there are sample labels, substitute them  
```{r}
if (samples_present) {
	for (index in seq_len(length(tree_list))) {
		tips <- tree_list[[index]]$tip.label
		new_tips <- tips
		for (tind in seq_len(length(tips))) {
			if (tips[tind] %in% sample_table$ID) {
				new_tips[tind] <- sample_table$label[sample_table$ID == tips[tind]]
			}
		}
		tree_list[[index]]$tip.label <- new_tips
	}
}
```


First, the concatenation tree  

Start a graphics device if wanting to output
```{r}
svglite("concat_out.svg", width = 28, height = 28)
```

Or pdf
```{r}
pdf("concat_out_gcf.pdf", width = 28, height = 42, family = "ArialMT")
```

Plot the concat tree with edge lengths
```{r}
tree <- tree_list[[1]]
plot.phylo(ladderize(tree, right = FALSE),
			no.margin = TRUE,
			font = 1)
add.scale.bar(x = mean(par("usr")[1:2]), y = par("usr")[3] + 1, font = 1)
```

OR, if wanting to plot in the reverse orientation
```{r}
tree <- tree_list[[1]]
plot.phylo(ladderize(tree, right = FALSE),
			direction = "leftwards",
			no.margin = TRUE,
			font = 1)
add.scale.bar(x = mean(par("usr")[1:2]), y = par("usr")[3] + 1, font = 1)
```

Plot node labels (UFbootstrap) on edges
```{r}
drawSupportOnEdges(tree$node.label, adj = c(0.5, -0.5), frame = "none")
```

Or, if wanting to plot node labels on nodes
```{r}
nodelabels(tree$node.label, adj = c(-0.05, 0.5), frame = "none")
```

Capture concordance factors to plot as desired  
Site concordance:  
```{r}
tree <- tree_list[[2]]
concord_df <- data.frame(matrix(ncol = 3, nrow = tree$Nnode))
index <- 1
for (label in tree$node.label) {
	if (any(label == "", label == "Root")) {
		values <- c("", "", "")
	} else {
		values <- strsplit(label, split = "\\/")[[1]]
	}
	concord_df[index, 1:3] <- values
	index <- index + 1
}
concord_df <- as.data.frame(lapply(concord_df, as.numeric))
```

Gene concordance:  
```{r}
tree <- tree_list[[3]]
gconcord_df <- data.frame(matrix(ncol = 4, nrow = tree$Nnode))
index <- 1
for (label in tree$node.label) {
	if (any(label == "", label == "Root")) {
		values <- c("", "", "", "")
	} else {
		values <- strsplit(label, split = "\\/")[[1]]
	}
	gconcord_df[index, 1:4] <- values
	index <- index + 1
}
gconcord_df <- as.data.frame(lapply(gconcord_df, as.numeric))
```

Plot concordance factors as pie charts on nodes 
Site concordance:  
```{r}
nodelabels(pie = concord_df,
			piecol = c("white", "grey", "black"),
			cex = 0.25)
```

Gene concordance:  
```{r}
nodelabels(pie = gconcord_df,
			piecol = c("white", "#c1c1c1", "#5b5a5a", "black"),
			cex = 0.25)
```

Stop creating the graphic
```{r}
invisible(dev.off())
```


Now for the ASTRAL tree  

Start a graphics device if wanting to output
```{r}
svglite("astral_out.svg", width = 22, height = 42)
```

Or pdf
```{r}
pdf("astral_out.pdf", width = 22, height = 42, family = "ArialMT")
```

Plot the tree with edge lengths
```{r}
tree <- tree_list[[4]]
plot.phylo(ladderize(tree, right = FALSE),
			no.margin = TRUE,
			font = 1)
add.scale.bar(x = mean(par("usr")[1:2]), y = par("usr")[3] + 1, font = 1)
```

Or, if wanting to plot in the reverse orientation
```{r}
tree <- tree_list[[4]]
plot.phylo(ladderize(tree, right = FALSE),
			direction = "leftwards",
			no.margin = TRUE,
			font = 1)
add.scale.bar(x = mean(par("usr")[1:2]), y = par("usr")[3] + 1, font = 1)
```

Grab the posterior probabilities from the file for display  
```{r}
pp_df <- data.frame(matrix(ncol = 1, nrow = tree$Nnode))
index <- 1
for (label in tree$node.label) {
	if (any(label == "", label == "Root")) {
		values <- c("")
	} else {
		values <- strsplit(label, split = "\\/")[[1]][1]
	}
	pp_df[index, ] <- values
	index <- index + 1
}
pp_df <- as.data.frame(lapply(pp_df, as.numeric))
pps <- sapply(pp_df[, 1], function(x) format(round(x, 2), nsmall = 2))
pps[pps == "NA"] <- ""
```

Plot node labels (posterior prob) on edges
```{r}
drawSupportOnEdges(pps, adj = c(0.5, -0.5), frame = "none")
```

Or, if wanting to plot node labels on nodes
```{r}
nodelabels(pps, adj = c(-0.05, 0.5), frame = "none")
```

Capture quartet support to plot  
```{r}
tree <- tree_list[[5]]
quart_df <- data.frame(matrix(ncol = 3, nrow = tree$Nnode))
index <- 1
for (label in tree$node.label) {
	if (any(label == "", label == "Root")) {
		values <- c("", "", "")
	} else {
		values <- strsplit(label, split = "\\/")[[1]]
	}
	quart_df[index, 1:3] <- values
	index <- index + 1
}
quart_df <- as.data.frame(lapply(quart_df, as.numeric))
```

Plot quartet support as pie charts on nodes
```{r}
nodelabels(pie = quart_df,
			piecol = c("white", "grey", "black"),
			cex = 0.25, adj = 0.49)
```

Stop creating the graphic
```{r}
invisible(dev.off())
```

If running the polytomy test, plot other trees
SVG
```{r}
if (run_poly) {
	svglite("astral_poly_out.svg", width = 22, height = 42)
	tree <- tree_list[[6]]
	plot.phylo(ladderize(tree, right = FALSE),
			no.margin = TRUE,
			use.edge.length = FALSE,
			node.depth = 2,
			font = 1,
			edge.width = 3)
	drawSupportOnEdges(tree$node.label, adj = c(0.5, -0.5), frame = "none")
	invisible(dev.off())
	svglite("astral_poly_collapsed_out.svg", width = 22, height = 42)
	tree <- tree_list[[7]]
	plot.phylo(ladderize(tree, right = FALSE),
			no.margin = TRUE,
			use.edge.length = FALSE,
			node.depth = 2,
			font = 1,
			edge.width = 3)
	drawSupportOnEdges(tree$node.label, adj = c(0.5, -0.5), frame = "none")
	invisible(dev.off())
}
```
Or pdf
```{r}
if (run_poly) {
	pdf("astral_poly_out.pdf", width = 22, height = 42, family = "ArialMT")
	tree <- tree_list[[6]]
	plot.phylo(ladderize(tree, right = FALSE),
			no.margin = TRUE,
			use.edge.length = FALSE,
			node.depth = 2,
			font = 1,
			edge.width = 3)
	drawSupportOnEdges(tree$node.label, adj = c(0.5, -0.5), frame = "none")
	invisible(dev.off())
	pdf("astral_poly_collapsed_out.pdf", width = 22, height = 42, family = "ArialMT")
	tree <- tree_list[[7]]
	plot.phylo(ladderize(tree, right = FALSE),
			no.margin = TRUE,
			use.edge.length = FALSE,
			node.depth = 2,
			font = 1,
			edge.width = 3)
	drawSupportOnEdges(tree$node.label, adj = c(0.5, -0.5), frame = "none")
	invisible(dev.off())
}
```