---
title: "Plot phylogenetic trees for sequence capture"
output: html_document
---

This document describes the process of plotting phylogenetic trees for sequence capture output
The trees need to be in Newick format and should be in the current directory

Results from IQ-TREE and ASTRAL will have multiple node labels (concordance factors)
To get the concordance trees into Newick properly, use `concord_to_newick.py`
Similarly, use `astral_parse.py` to separate node labels for ASTRAL (posterior probabilities and quartet support)

The outgroup file should have a single tip label per line
The correspondence between sample IDs and sample labels should be in a tab-separated text file, one line per sample
```{r}
suppressMessages(library(ape))
suppressMessages(library(phytools))
suppressMessages(library(svglite))
concat_tree <- "concat.treefile"
concat_scf <- "concord_newick_scf.tre"
concat_gcf <- "concord_newick_gcf.tre"
astral_p <- "astral_p.tre"
astral_q <- "astral_q.tre"
outgroup_file <- "outgroup.txt"
samples_file <- "samples.tab"
run_poly <- FALSE
```

If running a polytomy test, add that output
```{r}
run_poly <- TRUE
astral_t <- "astral_poly.tre"
astral_tc <- "astral_poly_collapsed.tre"
```

Read in the samples file
```{r}
if (file.exists(samples_file)) {
	samples_present <- TRUE
	sample_table <- read.table(samples_file, sep = "\t")
	colnames(sample_table) <- c("ID", "label")
} else {
	samples_present <- FALSE
}
```

Read in the outgroups file
```{r}
if (file.exists(outgroup_file)) {
	outgroup_present <- TRUE
	outgroups <- read.table(outgroup_file)[, 1]
} else {
	outgroup_present <- FALSE
}
```

Read in the tree files
```{r}
c_tree <- read.tree(concat_tree)
cs_tree <- read.tree(concat_scf)
cg_tree <- read.tree(concat_gcf)
ap_tree <- read.tree(astral_p)
aq_tree <- read.tree(astral_q)
if (run_poly) {
	at_tree <- read.tree(astral_t)
	atc_tree <- read.tree(astral_tc)
	tree_list <- list(c_tree, cs_tree, cg_tree, ap_tree, aq_tree, at_tree, atc_tree)
} else {
	tree_list <- list(c_tree, cs_tree, cg_tree, ap_tree, aq_tree)
}
```

If there is an outgroup, root the trees
```{r}
if (outgroup_present) {
	for (index in seq_len(length(tree_list))) {
		tree <- tree_list[[index]]
		these_outgroups <- outgroups[outgroups %in% tree$tip.label]
		rootnode <- getMRCA(tree, as.character(these_outgroups))
		position <- 0.5 * tree$edge.length[which(tree$edge[, 2] == rootnode)]
		rooted_tree <- reroot(tree, rootnode, position, edgelabel = TRUE)
		tree_list[[index]] <- rooted_tree
	}
}
```

For later creation of figures, it may be desirable to get the tip order of ladderized trees (before labels are substituted)  
This will output text files per tree, with sample IDs per line (order is from base of tree to top of page, i.e. starting from the outgroup)  
Based on: https://stackoverflow.com/a/34364914
```{r}
# we want tree 1 (concat), tree 4 (astral) and tree 7 (astral_poly_collapsed; if present)
tree_names <- c("concat", "concat_site", "concat_gene", "astral", "astral_q", "astral_poly", "astral_poly_collapsed")
if (run_poly) {
	indices <- c(1, 4, 7)
} else {
	indices <- c(1, 4)
}
for (index in indices) {
	lad_tree <- ladderize(tree_list[[index]], right = FALSE)
	# determine which edges are tips and get the order
	is_tip <- lad_tree$edge[, 2] <= length(lad_tree$tip.label)
	ordered_tips <- lad_tree$edge[is_tip, 2]
	# get the tips in order of plotting
	output_ordered_ids <- lad_tree$tip.label[ordered_tips]
	# write to a file
	connection <- file(paste0("tips_", tree_names[index], ".txt"))
	writeLines(output_ordered_ids, connection)
	close(connection)
}
```

If there are sample labels, substitute them
```{r}
if (samples_present) {
	for (index in seq_len(length(tree_list))) {
		tips <- tree_list[[index]]$tip.label
		new_tips <- tips
		for (tind in seq_len(length(tips))) {
			if (tips[tind] %in% sample_table$ID) {
				new_tips[tind] <- sample_table$label[sample_table$ID == tips[tind]]
			}
		}
		tree_list[[index]]$tip.label <- new_tips
	}
}
```

Determine likely graphics parameters for the pages needed to display the trees
(The page should probably not be smaller than about 8 wide x 11 tall)
```{r}
num_tips <- length(tree_list[[1]]$tip.label)
# for Arial 12 point, each letter is ~ 1/6 inch; we want to have a buffer of about 25%, so * 1.25
my_height <- num_tips * 0.21
if (my_height < 11) {
	my_height <- 11
}
# determine width so that it is narrower with more terminals
my_width <- my_height * (0.8 - 0.001 * num_tips)
# roughly set a "cex" value for displaying pie charts based on number of tips (and therefore width)
my_cex <- 0.65 - 0.002 * num_tips
```

It may be better to set these manually for your data and preference
```{r}
my_width <- 28
my_height <- 42
my_cex <- 0.25
```


**Plot trees**

First, the concatenation tree

Start a graphics device if wanting to output
(repeat if wanting a figure with, e.g., site concordance instead of gene concordance)
```{r}
pdf("concat_out_gcf.pdf", width = my_width, height = my_height, family = "ArialMT")
```

Plot the concat tree with edge lengths
```{r}
tree <- tree_list[[1]]
plot.phylo(ladderize(tree, right = FALSE),
	no.margin = TRUE,
	font = 1,
	edge.width = 2,
	label.offset = max(nodeHeights(tree)) / 200)
add.scale.bar(x = mean(par("usr")[1:2]),
	y = par("usr")[3] + 1,
	font = 1,
	lwd = 2)
```
Or, if wanting to plot in the reverse orientation
```{r}
tree <- tree_list[[1]]
plot.phylo(ladderize(tree, right = FALSE),
	direction = "leftwards",
	no.margin = TRUE,
	font = 1,
	edge.width = 2,
	label.offset = max(nodeHeights(tree)) / 200)
add.scale.bar(x = mean(par("usr")[1:2]),
	y = par("usr")[3] + 1,
	font = 1,
	lwd = 2)
```

Plot node labels (UFbootstrap) on edges
```{r}
drawSupportOnEdges(tree$node.label, adj = c(0.5, -0.5), frame = "none")
```
Or, if wanting to plot node labels on nodes
```{r}
nodelabels(tree$node.label, adj = c(-0.05, 0.5), frame = "none")
```

Capture concordance factors
```{r}
tree <- tree_list[[2]]
concord_df <- data.frame(matrix(ncol = 3, nrow = tree$Nnode))
index <- 1
for (label in tree$node.label) {
	if (any(label == "", label == "Root")) {
		values <- c("", "", "")
	} else {
		values <- strsplit(label, split = "\\/")[[1]]
	}
	concord_df[index, 1:3] <- values
	index <- index + 1
}
concord_df <- as.data.frame(lapply(concord_df, as.numeric))

tree <- tree_list[[3]]
gconcord_df <- data.frame(matrix(ncol = 4, nrow = tree$Nnode))
index <- 1
for (label in tree$node.label) {
	if (any(label == "", label == "Root")) {
		values <- c("", "", "", "")
	} else {
		values <- strsplit(label, split = "\\/")[[1]]
	}
	gconcord_df[index, 1:4] <- values
	index <- index + 1
}
gconcord_df <- as.data.frame(lapply(gconcord_df, as.numeric))
```

Plot as pie charts on nodes
Site:
```{r}
nodelabels(pie = concord_df,
	piecol = c("white", "grey", "black"),
	cex = my_cex)
```
Gene:
```{r}
nodelabels(pie = gconcord_df,
	piecol = c("white", "#c1c1c1", "#5b5a5a", "black"),
	cex = my_cex)
```

Stop creating the graphic
```{r}
invisible(dev.off())
```


Now for the ASTRAL tree

Start a graphics device if wanting to output
```{r}
pdf("astral_out.pdf", width = my_width, height = my_height, family = "ArialMT")
```

Plot the tree with edge lengths
```{r}
tree <- tree_list[[4]]
plot.phylo(ladderize(tree, right = FALSE),
	no.margin = TRUE,
	font = 1,
	edge.width = 2,
	label.offset = max(nodeHeights(tree)) / 200,
	x.lim =  max(nodeHeights(tree)) * 1.3)
add.scale.bar(x = mean(par("usr")[1:2]),
	y = par("usr")[3] + 1,
	font = 1,
	lwd = 2)
```
Or, if wanting to plot in the reverse orientation
```{r}
tree <- tree_list[[4]]
plot.phylo(ladderize(tree, right = FALSE),
	direction = "leftwards",
	no.margin = TRUE,
	font = 1,
	edge.width = 2,
	label.offset = max(nodeHeights(tree)) / 200,
	x.lim =  max(nodeHeights(tree)) * 1.3)
add.scale.bar(x = mean(par("usr")[1:2]),
	y = par("usr")[3] + 1,
	font = 1,
	lwd = 2)
```

Grab the posterior probabilities from the file for display
```{r}
pp_df <- data.frame(matrix(ncol = 1, nrow = tree$Nnode))
index <- 1
for (label in tree$node.label) {
	if (any(label == "", label == "Root")) {
		values <- c("")
	} else {
		values <- strsplit(label, split = "\\/")[[1]][1]
	}
	pp_df[index, ] <- values
	index <- index + 1
}
pp_df <- as.data.frame(lapply(pp_df, as.numeric))
pps <- sapply(pp_df[, 1], function(x) format(round(x, 2), nsmall = 2))
pps[pps == "NA"] <- ""
```

Plot node labels (posterior prob) on edges
```{r}
drawSupportOnEdges(pps, adj = c(0.5, -0.5), frame = "none")
```
Or, if wanting to plot node labels on nodes
```{r}
nodelabels(pps, adj = c(-0.05, 0.5), frame = "none")
```

Capture quartet support to plot
```{r}
tree <- tree_list[[5]]
quart_df <- data.frame(matrix(ncol = 3, nrow = tree$Nnode))
index <- 1
for (label in tree$node.label) {
	if (any(label == "", label == "Root")) {
		values <- c("", "", "")
	} else {
		values <- strsplit(label, split = "\\/")[[1]]
	}
	quart_df[index, 1:3] <- values
	index <- index + 1
}
quart_df <- as.data.frame(lapply(quart_df, as.numeric))
```

Plot quartet support as pie charts on nodes
```{r}
nodelabels(pie = quart_df,
	piecol = c("white", "grey", "black"),
	adj = 0.49,
	cex = my_cex)
```

Stop creating the graphic
```{r}
invisible(dev.off())
```


If running the polytomy test, plot other trees

The full tree with polytomy test results

Start creating the graphic
```{r}
pdf("astral_poly_out.pdf", width = my_width, height = my_height, family = "ArialMT")
```

Plot the tree
```{r}
tree <- tree_list[[6]]
plot.phylo(ladderize(tree, right = FALSE),
	no.margin = TRUE,
	use.edge.length = FALSE,
	node.depth = 2,
	font = 1,
	edge.width = 2,
	label.offset = 0.2)
```

Draw support
```{r}
drawSupportOnEdges(tree$node.label, adj = c(0.5, -0.5), frame = "none")
```

Stop creating the graphic
```{r}
invisible(dev.off())
```


The tree with branches failing the test collapsed

Start creating the graphic
```{r}
pdf("astral_poly_collapsed.pdf", width = my_width, height = my_height, family = "ArialMT")
```

Plot the tree
```{r}
tree <- tree_list[[7]]
plot.phylo(ladderize(tree, right = FALSE),
	no.margin = TRUE,
	use.edge.length = FALSE,
	node.depth = 2,
	font = 1,
	edge.width = 2,
	label.offset = 0.2)
```

Draw support (if desired)
```{r}
drawSupportOnEdges(tree$node.label, adj = c(0.5, -0.5), frame = "none")
```

Stop creating the graphic
```{r}
invisible(dev.off())
```
